Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-02-04 draft_lm1.tex
improved letter and a bit of the discussion

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
Our contribution is two-fold: (i) we show a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) we propose to 
use \rr to inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic.

@@

Since there is very limited information regarding the actual proportions of 
each virus in circulation and most information available is about the 
predominant serotypes for some epidemics in the period of 
study only~\cite{macedo_virological_2013}, we propose the use of a 
simplified a single strain model.
The main argument we put forward is that by conditioning on the \rr series, we 
implicitly take into account the variability introduced by the 
co-circulation of multiple serotypes and heterogeneous levels of immunity in 
the general population.
We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by health 
authorities (Oswaldo G. Cruz, personal communication).
Another source of uncertainty is under-reporting of diagnosed cases, which is a 
serious issue in the health care systems of many developing countries such as 
Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
It is interesting to notice that the larger epidemics, in terms of peak size are 
not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population in order to truly assess the impact of any epidemic. [WHY?]

>>>>

In this paper we propose a method to bypass the lack of serotype-specific case 
data by informing the time-varying force of infection with the instantaneous 
reproductive number, \rr which we calculate from aggregated data.
The main contribution of this paper can be summarized in the following items: 
(i) we show a method to quantify uncertainty 
about \rr that is readily applicable to other diseases and; (ii) we propose to 
use \rr to inform a dynamic epidemic model with time-varying force of infection 
in order to gain insight into the attack ratio of each epidemic; (iii) We 
propose an estimation procedure for circulating serotype's $S_0$ from 
aggregate case data, which is robust to epidemic sizes; We estimate the AR 
for 18 years of Dengue epidemics.

@@

Since there is very limited information regarding the actual proportions of 
each virus in circulation and most information available is about the 
predominant serotypes for some epidemics in the period of 
study only~\cite{macedo_virological_2013}, we propose the use of a 
simplified a single strain model.
The main argument we put forward is that by conditioning on the \rr series, we 
implicitly take into account the variability introduced by the 
co-circulation of multiple serotypes and heterogeneous levels of immunity in 
the general population.
We sought to deal with all important sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic, but not all could be satisfactorily 
addressed in this analysis.
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
It is estimated that for every case reported, up to 10-20 are not seen by health 
authorities~\cite{luz_disability_2009}.
Another source of uncertainty is under-reporting of diagnosed cases, which is a 
serious issue in the health care systems of many developing countries such as 
Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of Dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected Dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which are not easy to fully determine.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
It is interesting to notice that the larger epidemics, in terms of peak size are 
not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$.
